Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-01-26 paper-new/categorical-model.tex
updates mostly to extensional model

\section{Abstract Categorical Models of Graduality}\label{sec:abstract-models}

First, what is a categorical model of call-by-push-value? We will use
the following notion as our basic structure\footnote{There are models
that are closer to syntax e.g., which distinguish between value types
and contexts but this suffices for all of the models we consider}:
\begin{enumerate}
\item A cartesian category $\mathcal V$
\item A category $\mathcal E$
\item An action of $\mathcal V^{op}$ (with the $\mathcal V$ cartesian
  product as monoidal structure) on $\mathcal E$. We write this with
  an arrow $A \arr B$.

  This means we have 
  \[ \alpha : {A_1 \times A_2} \arr B \cong A_2 \arr (A_1 \arr B) \]
  and
  \[ i : 1 \arr B \cong B \]
  satisfying coherence isomorphisms.
\item A functor $U : \mathcal E \to \mathcal V$ that ``preserves
  powering'' in that every $U(A \arr B)$ is an exponential of $UB$ by $A$
  and that $U\alpha$ and $Ui$ are mapped to the canonical isomorphisms
  for exponentials.
\item A left adjoint $F \dashv U$
\end{enumerate}

\begin{example}
  Given a strong monad $T$ on a bicartesian closed category $\mathcal
  V$, we can extend this to a CBPV model by defining $\mathcal E$ to
  be the category $\mathcal V^T$ of algebras, defining $A \to B$ as
  the powering of algebras, $F$ as the free algebra and $U$ as the
  underlying object functor.
\end{example}

There are two relevant notions of \emph{morphism} of CBPV models:
\emph{strict} and \emph{lax}.
% morphisms of CBPV models
Given call-by-push-value models
$\mathcal M_1 = (\mathcal V_1, \mathcal E_1, \arr_1, U_1, F_1)$ and
$\mathcal M_2 = (\mathcal V_2, \mathcal E_2, \arr_2, U_2, F_2)$,
A \emph{strict} morphism $G$ from $M_1$ to $M_2$ is given by a pair of functors
$G_{\mathcal{V}}: V_1 \to V_2$ and $G_{\mathcal{E}} : E_1 \to E_2$
that strictly presere the constructors:
\begin{enumerate}
  \item $G_{\mathcal{E}} \circ F_1 = F_2 \circ G_{\mathcal{V}}$
  \item $G_{\mathcal{V}} \circ U_1 = U_2 \circ G_{\mathcal{E}}$
  \item $G_{\mathcal{E}}(A \arr_1 B) = G_{\mathcal{V}}(A) \arr_2 G_{\mathcal{E}}(B)$
  \item $G_{\mathcal{V}}(A_1 \times_1 A_2) = G_{\mathcal{V}}(A_1) \times_2 G_{\mathcal{V}}(A_2)$
  \item $G_{\mathcal{V}}(1_1) = 1_2$
\end{enumerate}
As well as strictly preserving the corresponding universal morphisms
and coherence isomorphisms.

A lax morphism instead preserves these only up to transformation
\begin{enumerate}
  \item $G_{\mathcal{E}} \circ F_1 \Rightarrow F_2 \circ G_{\mathcal{V}}$
  \item $G_{\mathcal{V}} \circ U_1 \Rightarrow U_2 \circ G_{\mathcal{E}}$
  \item $G_{\mathcal{E}}(A \arr_1 B) \Rightarrow G_{\mathcal{V}}(A) \arr_2 G_{\mathcal{E}}(B)$
  \item $G_{\mathcal{V}}(A_1 \times_1 A_2) \Rightarrow G_{\mathcal{V}}(A_1) \times_2 G_{\mathcal{V}}(A_2)$
  \item $G_{\mathcal{V}}(1_1) \Rightarrow 1_2$
\end{enumerate}
Additionally a lax morphism should have a relationship between these
transformations and the universal morphisms, but we will only consider
lax morphisms of thin categories, where such conditions hold
trivially.
% TODO what about the product and coproducts in V?
% Do we need that G_V(A \times_1 A') = G_V(A) \times_2 G_V(A')
% and likewise for coproducts?

With these definitions, we can define categories $\textrm{CBPV}_s$ and
$\textrm{CBPV}_l$ whose objects are CBPV models, and whose arrows are
CBPV morphisms.  This category will serve as the setting in which we
formulate the definition of a model of extensional gradual typing.

@@

An model $\mathcal{M}$ of extensional gradual typing consists of:
\begin{itemize}
  \item CBPV models $\mathcal M_f$ and $\mathcal M_{sq}$
  \item CBPV morphisms $r : \mathcal M_f \to \mathcal M_{sq}$ and 
  $s, t : \mathcal M_{sq} \to \mathcal M_f$
\end{itemize}

satisfying certain additional conditions that will be described below.

% % https://q.uiver.app/#q=WzAsMixbMCwxLCJcXG1hdGhjYWx7TX1fe3NxfSJdLFswLDAsIlxcbWF0aGNhbHtNfV9mIl0sWzEsMCwiciJdLFswLDEsInMiLDAseyJjdXJ2ZSI6LTJ9XSxbMCwxLCJ0IiwyLHsiY3VydmUiOjJ9XV0=
% \[\begin{tikzcd}[ampersand replacement=\&]
% 	{\mathcal{M}_f} \\
% 	{\mathcal{M}_{sq}}
% 	\arrow["r", from=1-1, to=2-1]
% 	\arrow["s", curve={height=-12pt}, from=2-1, to=1-1]
% 	\arrow["t"', curve={height=12pt}, from=2-1, to=1-1]
% \end{tikzcd}\]

Spelling this out in light of the above definitions, we see that this is
equivalent to the following in the category $\textbf{Cat}$:

% https://q.uiver.app/#q=WzAsNCxbMCwyLCJcXHZzcSJdLFsyLDIsIlxcZXNxIl0sWzAsMCwiXFx2ZiJdLFsyLDAsIlxcZWYiXSxbMiwzLCJcXEZmIiwwLHsiY3VydmUiOi0yfV0sWzMsMiwiXFxVZiIsMCx7ImN1cnZlIjotMn1dLFswLDEsIlxcRnNxIiwwLHsiY3VydmUiOi0yfV0sWzEsMCwiXFxVc3EiLDAseyJjdXJ2ZSI6LTJ9XSxbMiwwLCJcXHJ2Il0sWzAsMiwiXFxzdiIsMCx7ImN1cnZlIjotMn1dLFsyLDAsIlxcdHYiLDAseyJjdXJ2ZSI6LTJ9XSxbMSwzLCJcXHNlIiwwLHsiY3VydmUiOi0yfV0sWzMsMSwiXFx0ZSIsMCx7ImN1cnZlIjotMn1dLFszLDEsIlxccmUiXSxbNCw1LCJcXGJvdCIsMSx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFs2LDcsIlxcYm90IiwxLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
\[\begin{tikzcd}[ampersand replacement=\&]
	\vf \&\& \ef \\
	\\
	\vsq \&\& \esq
	\arrow[""{name=0, anchor=center, inner sep=0}, "\Ff", curve={height=-12pt}, from=1-1, to=1-3]
	\arrow[""{name=1, anchor=center, inner sep=0}, "\Uf", curve={height=-12pt}, from=1-3, to=1-1]
	\arrow[""{name=2, anchor=center, inner sep=0}, "\Fsq", curve={height=-12pt}, from=3-1, to=3-3]
	\arrow[""{name=3, anchor=center, inner sep=0}, "\Usq", curve={height=-12pt}, from=3-3, to=3-1]
	\arrow["\rv", from=1-1, to=3-1]
	\arrow["\sv", curve={height=-12pt}, from=3-1, to=1-1]
	\arrow["\tv", curve={height=-12pt}, from=1-1, to=3-1]
	\arrow["\se", curve={height=-12pt}, from=3-3, to=1-3]
	\arrow["\te", curve={height=-12pt}, from=1-3, to=3-3]
	\arrow["\re", from=1-3, to=3-3]
	\arrow["\bot"{description}, draw=none, from=0, to=1]
	\arrow["\bot"{description}, draw=none, from=2, to=3]
\end{tikzcd}\]

@@

That is, for the values, we have

\begin{enumerate}
  \item A cartesian category $\mathcal V_f$.
  The objects of $\mathcal V_f$ will be called \emph{value types}.
  The morphisms of $\mathcal V_f$ will be called \emph{(pure) functions}.

  \item A cartesian category $\mathcal V_{sq}$. 
  The objects of $\mathcal V_{sq}$ will be called \emph{value edges} or
  \emph{value relations}, and the morphisms are \emph{commuting squares}.

  \item Functors $\sv, \tv : \mathcal V_{sq} \to \mathcal V_f$ and 
  $\rv : \mathcal V_f \to \mathcal V_{sq}$.
\end{enumerate}

Likewise, we have the analogous definitions for computations.

We write $c : A \rel A'$ to mean that $c \in \ob(\vsq)$ such that 
$\sv(c) = A$ and $\tv(c) = A'$.
%
Likewise, let $c_i : A_i \rel A_i'$ and $c_o : A_o \rel A_o'$,
and let $f \in \vf(A_i, A_o)$ and $f' \in \vf(A_i', A_o')$.
The notation $\beta : f \ltdyn_{c_o}^{c_i} f'$ is defined to mean

\begin{enumerate}
  \item $\beta \in \vsq(c_i, c_o)$
  \item $\sv(\beta) = f$
  \item $\tv(\beta) = f'$
\end{enumerate}

(Recall that $\sv$ and $\tv$ are functors, so in addition to acting on
the objects of $\vsq$ they also act on morphisms.)

Picorially, this is depicted as a commuting square:

% https://q.uiver.app/#q=WzAsNCxbMCwwLCJBX2kiXSxbMSwwLCJBX2knIl0sWzAsMSwiQV9vIl0sWzEsMSwiQV9vJyJdLFswLDIsImYiLDJdLFsxLDMsImYnIl0sWzAsMSwiY19pIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMiwzLCJjX28iLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFs0LDUsIlxcYWxwaGEiLDEseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XV0=
\[\begin{tikzcd}[ampersand replacement=\&]
	{A_i} \& {A_i'} \\
	{A_o} \& {A_o'}
	\arrow[""{name=0, anchor=center, inner sep=0}, "f"', from=1-1, to=2-1]
	\arrow[""{name=1, anchor=center, inner sep=0}, "{f'}", from=1-2, to=2-2]
	\arrow["{c_i}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
	\arrow["{c_o}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
	\arrow["\alpha"{description}, draw=none, from=0, to=1]
\end{tikzcd}\]

% Note: When the identity of the square $\beta$ is not important, we may omit it
% and write $f \ltdyn_{c_o}^{c_i} f'$. In this case the meaning is that there exists
% a square $\beta : f \ltdyn_{c_o}^{c_i} f'$.

Composition of squares $\beta : f \ltdyn_{c_2}^{c_1} g$ and $\beta' : f' \ltdyn_{c_3}^{c_2} g'$
corresponds to ``stacking'' the square for $\beta'$ below the square for $\beta$.
Fuctoriality of $s$ and $t$ ensure that the left and right sides of the resulting square are as expected,
i.e., we get $\beta' \circ \beta : f' \circ f \ltdyn_{c_3}^{c_1} g' \circ g$.

% Fuctoriality of $s$ and $t$ ensure that we can ``vertically" compose 
% $\beta : f \ltdyn_{c_2}^{c_1} g$ and $\beta' : f' \ltdyn_{c_3}^{c_2} g'$
% to obtain $\beta' \circ \beta : f' \circ f \ltdyn_{c_3}^{c_1} g' \circ g$.
% Pictorially, this is represented by ``stacking'' the square for
% $\beta'$ below the square for $\beta$.

All of the above holds in an analogous manner for the computations.

We will work in ``locally thin'' models where there is at most one
square with a given boundary.
That is, if $\beta, \beta' : f \ltdyn_{c_o}^{c_i} f'$, then $\beta = \beta'$.
Thus, we may unambiguously omit the identity of the square, i.e., we may write
$f \ltdyn_{c_o}^{c_i} f'$.

We also have a ``horizontal" composition operation on value edges and on computation edges.
That is, let 
%
\[ X = \{ (c, c') \in \ob(\vsq) \times \ob(\vsq) \mid \tv(c) = \sv(c') \}. \]
%
There is an operation $\comp : X \to \ob(\vsq)$ such that $\sv(c \comp c') = \sv(c)$
and $\tv(c \comp c') = \tv(c')$. Likewise for computations.
%
Importantly, we emphasize that this composition is \emph{not} a functor: we only
require that it act on \emph{objects} of $\vsq$ (i.e. edges) and not on the morphisms
(i.e. the squares). Intuivitely, this has to do with the fact that in the extensional
setting, the semantic term precision function is \emph{not} transitive.

With this definition, is easily shown that there is a category $\ve$ of value relations,
whose objects are the objects of $\vf$, and such that $\ve(A, A')$ is the set of objects
$c$ of $\vsq$ whose source is $A$ and whose target is $A'$.
Composition of morphisms is defined using the above operation $\comp$.
Similarly, we have a category $\ee$ of computation relations.

We also require that the category $\ve$ of relations is thin ``up to an
identity square'', i.e., for any $c, c' \in \ve(A, A')$ we have that the
following square commutes:

\[\begin{tikzcd}[ampersand replacement=\&]
  A \& {A'} \\
  A \& {A'}
  \arrow[from=1-1, to=2-1, Rightarrow, no head]
  \arrow[from=1-2, to=2-2, Rightarrow, no head]
  \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
  \arrow["c'"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
\end{tikzcd}\]

@@

Finally, we specify the abstract behavior of casts.
In brief, given $c : A \rel A'$, an upcast will be a morphism $\upf\, c$ in $\vf(A, A')$
such that $\upf\, c$ ``represents $c$" (we will define this shortly). % such that $c$ is \emph{left-representable} by $\up c$.
In reality, it is slightly more complicated, since we need to build composition of edges
into the definition, but the idea is similar.

In particular, let $c : A \rel A'$ and $f \in \mathcal V_f(A, A')$. We
say that $c$ is \emph{left-representable by} $f$ if for any $c_l : A_l
\rel A$ and $c_r : A' \rel A_r$ the following squares commute\max{is ``commute'' good terminology?}:
\begin{center}
  \begin{tabular}{ m{14em} m{14em} } 
    % UpL
    \begin{tikzcd}[ampersand replacement=\&]
      A \& {A'} \& {A_r} \\
      {A'} \&\& {A_r}
      \arrow["f"', from=1-1, to=2-1]
      %
      \arrow[from=1-3, to=2-3, Rightarrow, no head]
      %
      \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
      \arrow["c_r", "\shortmid"{marking}, no head, from=1-2, to=1-3]
      %
      \arrow["c_r"', "\shortmid"{marking}, no head, from=2-1, to=2-3]
    \end{tikzcd}
    &
    % UpR
    \begin{tikzcd}[ampersand replacement=\&]
      A_l \&\& {A} \\
      {A_l} \& {A} \& {A'}
      \arrow[from=1-1, to=2-1, Rightarrow, no head]
      %
      \arrow["f", from=1-3, to=2-3]
      %
      \arrow["c_l", "\shortmid"{marking}, no head, from=1-1, to=1-3]
      %
      \arrow["c_l"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
      \arrow["c"', "\shortmid"{marking}, no head, from=2-2, to=2-3]
    \end{tikzcd}
  \end{tabular}
\end{center}

Dually, let $d \in B \rel B'$ and $\phi \in \mathcal E_f(B', B)$.  We
say that $d$ is \emph{right-representable by} $\phi$ if for any $d_l :
B_l \rel B$ and $d_r : B' \rel B_r$ the following two squares commute:

\begin{center}
  \begin{tabular}{ m{14em} m{14em} } 
    % DnR
    \begin{tikzcd}[ampersand replacement=\&]
      {B_l} \& {B} \& {B'} \\
      {B_l} \&\& {B}
      \arrow[from=1-1, to=2-1, Rightarrow, no head]
      %
      \arrow["\phi", from=1-3, to=2-3]
      %
      \arrow["d_l", "\shortmid"{marking}, no head, from=1-1, to=1-2]
      \arrow["d", "\shortmid"{marking}, no head, from=1-2, to=1-3]
      %
      \arrow["d_l"', from=2-1, to=2-3, no head]
    \end{tikzcd}
    &
    % DnL
    \begin{tikzcd}[ampersand replacement=\&]
      {B'} \&\& {B_r} \\
      {B} \& {B'} \& {B_r}
      \arrow["\phi"', from=1-1, to=2-1]
      %
      \arrow[from=1-3, to=2-3, Rightarrow, no head]
      %
      \arrow["d_r", from=1-1, to=1-3, no head]
      %
      \arrow["d"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
      \arrow["d_r"', "\shortmid"{marking}, no head, from=2-2, to=2-3]
    \end{tikzcd}
  \end{tabular}
\end{center}

Then we formulate the relationship between relation morphisms and
function morphisms as follows:
\begin{enumerate}
\item There is an identity-on-objects functor $\upf : \mathcal V_e \to
  \mathcal V_f$ such that every $c$ is left-representable by $\upf(c)$.
\item There is an identity-on-objects functor $\dnf : \mathcal
  E_e^{op} \to \mathcal \mathcal E_f$ such that every $d$ is
  right-representable by $\dnf(d)$.
\end{enumerate}

@@

In summary, an extensional model consists of:
\begin{enumerate}
  \item A reflexive graph internal to CBPV models.
  \item Composition of value and computation relations that form a category with the reflexive relations as identity.
  \item An identity-on-objects functor $\upf : \mathcal V_e \to \mathcal V_f$ taking each value relation to a morphism that left-represents it.
  \item An identity-on-objects functor $\dnf : \mathcal E_e^{op} \to \mathcal E_f$ taking each computation relation to a morphism that right-represents it.
  \item The CBPV connectives $U,F,\times,\to$ are all \emph{covariant} functorial on relations\footnote{the reflexive graph structure already requires that these functors preserve identity relations}
    \begin{itemize}
    \item $U(dd') = U(d)U(d')$
    \item $F(cc') = F(c)F(c')$
    \item $(cc') \to (dd') = (c \to d)(c' \to d')$
    \item $(c_1c_1') \times (c_2c_2') = (c_1 \times c_2)(c_1'\times c_2')$
    \end{itemize}
  \item A distinguished value type $D$ with distinguished relation
    $\injarr{}: U(D \to F D) \rel D$ satisfying $\dnc {\injarr{}}F(\upc{\injarr{}}) = \id$
\end{enumerate}

\subsection{Stuff about Intensional Models}

>>>>

\section{Revised Categorical Models of Graduality}\label{sec:abstract-models}

Next, we develop our appropriate weakened notions of models of
graduality, which we divide into \emph{extensional} models, where
ordering is up to weak bisimilarity, and \emph{intensional} models,
where ordering relates terms that considers computational steps to be
observable. We develop the notion of intensional model of gradual
typing in stages and show how to develop one from a base model of
effectful functions and relations.

@@

Since we lack transitivity of ordering when reasoning in guarded type
theory, our weakened notion of extensional model is based on reflexive
graph categories rather than double categories. This means we lose the
reasoning principle of horizontal pasting of squares. We will still
require a notion of \emph{composition} of relations, to model the
transitivity of type precision. We note that without horizontal
pasting of squares, the notion of left/right representability of
squares is not sufficient to interpret the cast rules of gradual
typing. Instead we generalize it slightly as follows.

Let $c : A \rel A'$ and $f : A \to A'$ in a reflexive graph category
with composition of relations. We say that $c$ is \emph{universally
left-representable by} $f$ if for any $c_l : A_l \rel A$ and $c_r : A'
\rel A_r$ we have $f \ltsq{cc_r}{c_r} \id$ and $\id \ltsq{c_l}{c_lc}
f$. Dually, let $d : B \rel B'$ and $g : B' \to B$.  We say that $d$
is \emph{universally right-representable by} $g$ if for any $d_l : B_l
\rel B$ and $d_r : B' \rel B_r$ we have $\id \ltsq{d_ld}{d_l} \phi$
and $\phi \ltsq{d_r}{dd_r}\id$. In a reflexive graph category these
are stronger than the definitions of left/right representability since
we can pick $c_l,c_r,d_l,d_r$ to be the reflexive relations. In a
double category these are equivalent, but the equivalence uses
horizontal pasting.

Additionally, while in the presence of horizontal pasting
compositionality of representable morphisms is automatic, without this
principle we must require it explicitly. Finally, we explicitly add in
a requirement that the type constructors are functorial in the

In summary, an extensional model consists of:
\begin{enumerate}
  \item A locally thin reflexive graph internal to CBPV models.
  \item Composition of value and computation relations that form a category with the reflexive relations as identity.
  \item An identity-on-objects functor $\upf : \mathcal V_e \to \mathcal V_f$ taking each value relation to a morphism that universally left-represents it.
  \item An identity-on-objects functor $\dnf : \mathcal E_e^{op} \to \mathcal E_f$ taking each computation relation to a morphism that universally right-represents it.
  \item The CBPV connectives $U,F,\times,\to$ are all \emph{covariant} functorial on relations\footnote{the reflexive graph structure already requires that these functors preserve identity relations}
    \begin{itemize}
    \item $U(dd') \equidyn U(d)U(d')$
    \item $F(cc') \equidyn F(c)F(c')$
    \item $(cc') \to (dd') \equidyn (c \to d)(c' \to d')$
    \item $(c_1c_1') \times (c_2c_2') \equidyn (c_1 \times c_2)(c_1'\times c_2')$
    \end{itemize}
    where $c \equidyn c'$ means $\id \ltsq{c}{c'}\id$ and $\id \ltsq{c'}{c} \id$.
  \item Distinguished value type $\nat$ with morphisms $z : \mathcal
    V(1,\nat)$ and $s : \mathcal V(\nat,\nat)$.
  \item Distinguished value types $D$ with distinguished relations
    $\iarr{}: U(D \to F D) \rel D$ and $\inat : \nat \rel D$ and $\itimes : D \times D \rel D$
    each satisfying the retraction property $\dnc {\injarr{}}F(\upc{\injarr{}}) \equidyn \id$.
\end{enumerate}

%% %% A model $\mathcal{M}$ of extensional gradual typing consists of the
%% %% following:
%% \begin{itemize}
%% \item A reflexive graph internal to the category of CBPV models and
%%   strict morphisms that is locally thin: there is at most one morphism
%%   in $\mathcal V_{sq}$ for any given domain, codomain, source and
%%   target, and as well for $\mathcal E$.
%% \item 
%% \item A natural transformation $\mho : 1 \Rightarrow U$ satisfying $\mho \circ ! \ltsq{r(A)}{r(B)} f$ for every $f$.
%% \item Such that all relation value morphisms are \emph{universally}
%%   left representable and all relation computation morphisms are
%%   universally right representable.
%% \end{itemize}

%% satisfying certain additional conditions that will be described below.

%% % % https://q.uiver.app/#q=WzAsMixbMCwxLCJcXG1hdGhjYWx7TX1fe3NxfSJdLFswLDAsIlxcbWF0aGNhbHtNfV9mIl0sWzEsMCwiciJdLFswLDEsInMiLDAseyJjdXJ2ZSI6LTJ9XSxbMCwxLCJ0IiwyLHsiY3VydmUiOjJ9XV0=
%% % \[\begin{tikzcd}[ampersand replacement=\&]
%% % 	{\mathcal{M}_f} \\
%% % 	{\mathcal{M}_{sq}}
%% % 	\arrow["r", from=1-1, to=2-1]
%% % 	\arrow["s", curve={height=-12pt}, from=2-1, to=1-1]
%% % 	\arrow["t"', curve={height=12pt}, from=2-1, to=1-1]
%% % \end{tikzcd}\]

%% Spelling this out in light of the above definitions, we see that this is
%% equivalent to the following in the category $\textbf{Cat}$:

%% % https://q.uiver.app/#q=WzAsNCxbMCwyLCJcXHZzcSJdLFsyLDIsIlxcZXNxIl0sWzAsMCwiXFx2ZiJdLFsyLDAsIlxcZWYiXSxbMiwzLCJcXEZmIiwwLHsiY3VydmUiOi0yfV0sWzMsMiwiXFxVZiIsMCx7ImN1cnZlIjotMn1dLFswLDEsIlxcRnNxIiwwLHsiY3VydmUiOi0yfV0sWzEsMCwiXFxVc3EiLDAseyJjdXJ2ZSI6LTJ9XSxbMiwwLCJcXHJ2Il0sWzAsMiwiXFxzdiIsMCx7ImN1cnZlIjotMn1dLFsyLDAsIlxcdHYiLDAseyJjdXJ2ZSI6LTJ9XSxbMSwzLCJcXHNlIiwwLHsiY3VydmUiOi0yfV0sWzMsMSwiXFx0ZSIsMCx7ImN1cnZlIjotMn1dLFszLDEsIlxccmUiXSxbNCw1LCJcXGJvdCIsMSx7InNob3J0ZW4iOnsic291cmNlIjoyMCwidGFyZ2V0IjoyMH0sInN0eWxlIjp7ImJvZHkiOnsibmFtZSI6Im5vbmUifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFs2LDcsIlxcYm90IiwxLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
%% \[\begin{tikzcd}[ampersand replacement=\&]
%% 	\vf \&\& \ef \\
%% 	\\
%% 	\vsq \&\& \esq
%% 	\arrow[""{name=0, anchor=center, inner sep=0}, "\Ff", curve={height=-12pt}, from=1-1, to=1-3]
%% 	\arrow[""{name=1, anchor=center, inner sep=0}, "\Uf", curve={height=-12pt}, from=1-3, to=1-1]
%% 	\arrow[""{name=2, anchor=center, inner sep=0}, "\Fsq", curve={height=-12pt}, from=3-1, to=3-3]
%% 	\arrow[""{name=3, anchor=center, inner sep=0}, "\Usq", curve={height=-12pt}, from=3-3, to=3-1]
%% 	\arrow["\rv", from=1-1, to=3-1]
%% 	\arrow["\sv", curve={height=-12pt}, from=3-1, to=1-1]
%% 	\arrow["\tv", curve={height=-12pt}, from=1-1, to=3-1]
%% 	\arrow["\se", curve={height=-12pt}, from=3-3, to=1-3]
%% 	\arrow["\te", curve={height=-12pt}, from=1-3, to=3-3]
%% 	\arrow["\re", from=1-3, to=3-3]
%% 	\arrow["\bot"{description}, draw=none, from=0, to=1]
%% 	\arrow["\bot"{description}, draw=none, from=2, to=3]
%% \end{tikzcd}\]

@@

%% That is, for the values, we have

%% \begin{enumerate}
%%   \item A cartesian category $\mathcal V_f$.
%%   The objects of $\mathcal V_f$ will be called \emph{value types}.
%%   The morphisms of $\mathcal V_f$ will be called \emph{(pure) functions}.

%%   \item A cartesian category $\mathcal V_{sq}$. 
%%   The objects of $\mathcal V_{sq}$ will be called \emph{value edges} or
%%   \emph{value relations}, and the morphisms are \emph{commuting squares}.

%%   \item Functors $\sv, \tv : \mathcal V_{sq} \to \mathcal V_f$ and 
%%   $\rv : \mathcal V_f \to \mathcal V_{sq}$.
%% \end{enumerate}

%% Likewise, we have the analogous definitions for computations.

%% We write $c : A \rel A'$ to mean that $c \in \ob(\vsq)$ such that 
%% $\sv(c) = A$ and $\tv(c) = A'$.
%% %
%% Likewise, let $c_i : A_i \rel A_i'$ and $c_o : A_o \rel A_o'$,
%% and let $f \in \vf(A_i, A_o)$ and $f' \in \vf(A_i', A_o')$.
%% The notation $\beta : f \ltdyn_{c_o}^{c_i} f'$ is defined to mean

%% \begin{enumerate}
%%   \item $\beta \in \vsq(c_i, c_o)$
%%   \item $\sv(\beta) = f$
%%   \item $\tv(\beta) = f'$
%% \end{enumerate}

%% (Recall that $\sv$ and $\tv$ are functors, so in addition to acting on
%% the objects of $\vsq$ they also act on morphisms.)

%% Picorially, this is depicted as a commuting square:

%% % https://q.uiver.app/#q=WzAsNCxbMCwwLCJBX2kiXSxbMSwwLCJBX2knIl0sWzAsMSwiQV9vIl0sWzEsMSwiQV9vJyJdLFswLDIsImYiLDJdLFsxLDMsImYnIl0sWzAsMSwiY19pIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMiwzLCJjX28iLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFs0LDUsIlxcYWxwaGEiLDEseyJzaG9ydGVuIjp7InNvdXJjZSI6MjAsInRhcmdldCI6MjB9LCJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJub25lIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XV0=
%% \[\begin{tikzcd}[ampersand replacement=\&]
%% 	{A_i} \& {A_i'} \\
%% 	{A_o} \& {A_o'}
%% 	\arrow[""{name=0, anchor=center, inner sep=0}, "f"', from=1-1, to=2-1]
%% 	\arrow[""{name=1, anchor=center, inner sep=0}, "{f'}", from=1-2, to=2-2]
%% 	\arrow["{c_i}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
%% 	\arrow["{c_o}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
%% 	\arrow["\alpha"{description}, draw=none, from=0, to=1]
%% \end{tikzcd}\]

%% % Note: When the identity of the square $\beta$ is not important, we may omit it
%% % and write $f \ltdyn_{c_o}^{c_i} f'$. In this case the meaning is that there exists
%% % a square $\beta : f \ltdyn_{c_o}^{c_i} f'$.

%% Composition of squares $\beta : f \ltdyn_{c_2}^{c_1} g$ and $\beta' : f' \ltdyn_{c_3}^{c_2} g'$
%% corresponds to ``stacking'' the square for $\beta'$ below the square for $\beta$.
%% Fuctoriality of $s$ and $t$ ensure that the left and right sides of the resulting square are as expected,
%% i.e., we get $\beta' \circ \beta : f' \circ f \ltdyn_{c_3}^{c_1} g' \circ g$.

%% % Fuctoriality of $s$ and $t$ ensure that we can ``vertically" compose 
%% % $\beta : f \ltdyn_{c_2}^{c_1} g$ and $\beta' : f' \ltdyn_{c_3}^{c_2} g'$
%% % to obtain $\beta' \circ \beta : f' \circ f \ltdyn_{c_3}^{c_1} g' \circ g$.
%% % Pictorially, this is represented by ``stacking'' the square for
%% % $\beta'$ below the square for $\beta$.

%% All of the above holds in an analogous manner for the computations.

%% We will work in ``locally thin'' models where there is at most one
%% square with a given boundary.
%% That is, if $\beta, \beta' : f \ltdyn_{c_o}^{c_i} f'$, then $\beta = \beta'$.
%% Thus, we may unambiguously omit the identity of the square, i.e., we may write
%% $f \ltdyn_{c_o}^{c_i} f'$.

%% We also have a ``horizontal" composition operation on value edges and on computation edges.
%% That is, let 
%% %
%% \[ X = \{ (c, c') \in \ob(\vsq) \times \ob(\vsq) \mid \tv(c) = \sv(c') \}. \]
%% %
%% There is an operation $\comp : X \to \ob(\vsq)$ such that $\sv(c \comp c') = \sv(c)$
%% and $\tv(c \comp c') = \tv(c')$. Likewise for computations.
%% %
%% Importantly, we emphasize that this composition is \emph{not} a functor: we only
%% require that it act on \emph{objects} of $\vsq$ (i.e. edges) and not on the morphisms
%% (i.e. the squares). Intuivitely, this has to do with the fact that in the extensional
%% setting, the semantic term precision function is \emph{not} transitive.

%% With this definition, is easily shown that there is a category $\ve$ of value relations,
%% whose objects are the objects of $\vf$, and such that $\ve(A, A')$ is the set of objects
%% $c$ of $\vsq$ whose source is $A$ and whose target is $A'$.
%% Composition of morphisms is defined using the above operation $\comp$.
%% Similarly, we have a category $\ee$ of computation relations.

%% We also require that the category $\ve$ of relations is thin ``up to an
%% identity square'', i.e., for any $c, c' \in \ve(A, A')$ we have that the
%% following square commutes:

%% \[\begin{tikzcd}[ampersand replacement=\&]
%%   A \& {A'} \\
%%   A \& {A'}
%%   \arrow[from=1-1, to=2-1, Rightarrow, no head]
%%   \arrow[from=1-2, to=2-2, Rightarrow, no head]
%%   \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
%%   \arrow["c'"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
%% \end{tikzcd}\]

@@

%% Then we formulate the relationship between relation morphisms and
%% function morphisms as follows:
%% \begin{enumerate}
%% \item There is an identity-on-objects functor $\upf : \mathcal V_e \to
%%   \mathcal V_f$ such that every $c$ is left-representable by $\upf(c)$.
%% \item There is an identity-on-objects functor $\dnf : \mathcal
%%   E_e^{op} \to \mathcal \mathcal E_f$ such that every $d$ is
%%   right-representable by $\dnf(d)$.
%% \end{enumerate}

@@
